Modular Model-Checking for Fault-Tolerant Distributed System Implementations

### 2016-12-07 doc/om-paper/paper.tex
Add model sketch section, futz with experiments table, add Rushby benchmark data

\usepackage{graphicx}
\usepackage{alltt}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

@@

\newcommand{\OM}[1]{\ensuremath{\mathrm{OM}(#1)}}
\newcommand{\OMH}{\ensuremath{\mathrm{OMH}}}
\newcommand{\Id}{\ensuremath{\mathrm{Id}}}
\newcommand{\Msg}{\ensuremath{\mathrm{Msg}}}

@@

In Section~\ref{sec:experimental}, we give describe two sets of experiments on the model. First, we benchmark the model, showing that despite the additional complexity, it scales farther than other model-checking verifications at the algorithmic level. The scalability is due principally to specifying invariants and using $k$-induction in SRI's SAL model-checker~\cite{sal}. Developing invariants requires some user guidance. A user should not have to discover all new invariants for new protocols or implementations. We show that most of the lemmas required are not protocol specific and are modular, concerning only specific aspects of the system, such as the timing, faults, or local node behavior. To demonstrate this, we modify the OHM implementation and show that in each case, only a small number of localized modifications are required to repair the model and proofs. The modifications we study are to add omissive-asymetric faults, make the implementation time-triggered, and replace the majority vote with a mid-value selection.

@@

First, we briefly describe the algorithm, sketch our model of the algorithm within the model presented in Section~\ref{sec:model}, then describe it's invariants.

\ben{The word "model" is very overloaded in the paper so far. I wonder if we could use a different word like "modeling framework", or something more concise, for the one conotation and just "model" for the concrete mathematical description of a particular system?}

\subsection{$\OMH(m)$ Algorithm}
$\OMH$ is a recursive algorithm that proceeds in rounds of communication. Let $\OMH(m)$ be a recursive function implementing the algorithm, parameterized by the number of rounds, $m$. Consider a finite set of nodes $N$. Distinguish one node as the \emph{general}, $g$, and the remaining nodes $L = N \setminus \{g\}$ as the \emph{lieutenants}. We assume the identity of any general or lieutenant cannot be spoofed. Broadcast communication proceeds in rounds. Denote any message that is detectably faulty (e.g., fails a CRC) or is absent, by $ERR$. Additionally, in the algorithm, nodes report on values they have previously received. In doing so, nodes must differentiate \emph{reporting} $ERR$ from an $ERR$ itself. Let $R$ denote that an error is being reported. Finally, let $V$ be a special, designated value.

@@

\begin{itemize}
\item {\bf $OM(0)$}: $g$ broadcasts a value to each lieutenant, and the lieutenants return the value received (or $ERR$).
\item {\bf $OM(m)$, $m > 0$}:
  \begin{enumerate}
  \item $g$ broadcasts a value to each lieutenant, $l$.
  \item\label{om:two} Let $l_v$ be the value received by $l \in L$ from $g$. Then for each $l$, execute $\OMH(m-1)$, assigning $l$ to be the general and $L \setminus \{l\}$ to be the lieutenants. $l$ sends $l_v$ or $R$, if $l_v = ERR$.
  \item For each lieutenant $l \in L$, remove all $ERR$ values received in Step~\ref{om:two} from executing $OHM(m-1)$. Compute the majority value over the remaining values, or $V$ if there is no majority. If the majority value is $R$, return $E$.
  \end{enumerate}
\end{itemize}

@@

$\OMH$ is designed to preserve \emph{validity} and \emph{agreement}
properties. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. More formally, let the general broadcast $v \neq ERR$, and let $l_i$ denote the output for lieutenant $i$:
%
\begin{equation}
  \tag{Validity}
    \forall \,i. \quad l_i = v
\end{equation}
%
Agreement states that each lieutenant outputs the same value. Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively:
%
\begin{equation}
  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j
\end{equation}
%

@@

\subsection{Model Sketch}\label{sec:sketch}
\lee{Ben: can you introduce the relay/receiver unrolling of lieutenants here, since I refer to them later?}
\lee{explain why there are 7 transition systems defined (source, recv, relay, abstractor, abstractorMonitor, clock, observer }

The verification is carried out in the Symbolic Analysis Laboratory~\cite{}, which contains a suite of model-checkers and a high-level specification language. In our work, we use infinite-state (SMT-based) $k$-induction~\cite{}.

\lee{Maybe talk about MJRTY here. Maybe also talk about relaxing in synchronous communication in model.}
\lee{talk about synchrony being loosened, so the general/receivers don't all send/receive at the same time. Talk about lieutenants being ``unrolled'' into relays and receivers, following Rushby.}

@@

\begin{figure}
  \centering
  \begin{tabular}{r||c|c|c|c|c|c|c|c|c|c|}
      \multicolumn{11}{c}{Receivers} \\
      \cline{2-11}
   Relays &   1  &  2  &   3  &  4  &   5  &  6   &  7   &  8  &  9  &  10 \\
      \hline \hline
      1   & 7    & 9   & 12   & 15  & 21   & 25   & 32   & 40  & 54  & 74  \\
      \hline
      2   & 17   & 14  & 21   & 30  & 42   & 53   & 74   & 99  & 144 & -   \\
      \hline
      3   & 21   & 22  & 40   & 50  & 81   & 102  & 155  & 279 & -   & -   \\
      \hline
      4   & 27   & 34  & 59   & 99  & 141  & 237  & 1114 & -   & -   & -   \\
      \hline
      5   & 22   & 94  & 125  & 335 & TO   & 1406 & -    & -   & -   & -   \\
      \hline
      6   & 36   & 132 & 2966 & 844 & 2457 & -    & -    & -   & -   & -   \\
      \hline
      7   & 83   & 487 & TO   & TO  & -    & -    & -    & -   & -   & -   \\
      \hline
      8   & 298  & TO  & TO   & -   & -    & -    & -    & -   & -   & -   \\
      \hline
      9   & 1428 & TO  & -    & -   & -    & -    & -    & -   & -   & -   \\
      \hline
     10   & TO   & -   & -    & -   & -    & -    & -    & -   & -   & -   \\
      \hline
  \end{tabular}
  \label{fig:benchmark}
  \caption{Benchmark of full proof computation time for $\OMH(1)$ implementation. Times are in seconds with a timeout (TO) limit of one hour. Dashes ('-') denote no benchmark was run.}
\end{figure}

@@

\lee{Ben: can you run a couple of representative examples of Rushby's model, say for 5 and 6 receivers?}

@@

\begin{figure}
  \centering
  \begin{tabular}{|p{2.25cm}|p{2.25cm}|p{2.25cm}|p{2.25cm}|p{2.25cm}|}
  \hline
                                  & Transition systems (7 total)  &Definitions\linebreak (58 total) & Invariants\linebreak (11 total) & Invariant classes (5 total) \\
  \hline \hline
  No Fault Model                  & receivers & 8 modified & 2 modified & ASM, faults \\
  \hline
  Omissive-asymetric faults       & None      & 4 modified & 2 modified & faults             \\
  \hline
  Time-triggered messaging        & source, relays, receivers & 3 new & 2 new, 3 modified & calendar, faults \\
  \hline
  Mid-value selection             & receivers & 4 new, 3 modified & 2 modified & ASM, voting \\
  \hline
  \end{tabular}
  \caption{Refactoring effort for protocol modifications, measured by which portions of the model have to be modified.}
  \label{fig:effort}
\end{figure}

@@

\paragraph{Omissive asymmetric faults.}
Removing faults already described by the fault model is easy. Recall that in our model faults do not appear in the system specification and only operate on the calendar. Removing a fault from the system requires only setting the number of a particular kind of faults to zero in the maximum fault assumption.

@@

\paragraph{Time-triggered messaging.}
A \emph{time-triggered} distributed system is one in which nodes are independently clocked, but clock constraints (ensured by periodic clock synchronization) allow the model to appear as if it is executing synchronously~\cite{}.

@@

\paragraph{Mid-value selection.}
Our $\OMH(1)$ model leverages majority vote in order to tolerate faults. Another choice for the fault masking algorithm used here is mid-value select (or median). This choice is more common in applications involving hardware, signal selection (where signal values cannot be relied on to exactly match even in the absence of faults), or cases where information about congruence is useful. To implement mid-value select in our model, we allow messages sent to take values in $\mathbb{R}$ and the receiver transition system is modified in two ways. First, a second buffer is introduced which will hold the sorted contents of the main buffer once voting has commenced. Second, a mid-value select function is called on the sorted buffer and the result is stored as the receiver's vote. The only invariants needing modification were the ASM definition (to account for the values stored by the new buffer and the relation between it and the main buffer) and the voting invariant.

\subsection{Proof Effort Remarks}
The lemmas described in Section~\ref{sec:invariants} are constructed by-hand and represent multiple days of effort, but that effort includes both model and protocol construction and generalization as well as verification. The counterexamples returned by SAL are very useful for strengthening invariants, but tedious to analyze---a model with five relays and two receivers contains 90 state variables, and there are known counterexamples to models that size~\cite{csl-93-2}. Once we developed the ASM synchronously-composed observer, the verification effort was sped up considerably.

>>>>

\usepackage{graphicx}
\usepackage{alltt}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{multirow}

@@

\newcommand{\OM}[1]{\ensuremath{\mathrm{OM}(#1)}}
\newcommand{\OMH}{\ensuremath{\mathrm{OMH}}}
\newcommand{\Id}{\ensuremath{\mathrm{Id}}}
\newcommand{\Msg}{\ensuremath{\mathrm{Msg}}}
\newcommand{\ERR}{\ensuremath{\mathrm{ERR}}}
\newcommand{\TO}{\ensuremath{\bm{\mathrm{T}}}}

@@

In Section~\ref{sec:experimental}, we give describe two sets of experiments on the model. First, we benchmark the model, showing that despite the additional complexity, it scales farther than other model-checking verifications at the algorithmic level. The scalability is due principally to specifying invariants and using $k$-induction in SRI's SAL model-checker~\cite{SAL}. Developing invariants requires some user guidance. A user should not have to discover all new invariants for new protocols or implementations. We show that most of the lemmas required are not protocol specific and are modular, concerning only specific aspects of the system, such as the timing, faults, or local node behavior. To demonstrate this, we modify the OHM implementation and show that in each case, only a small number of localized modifications are required to repair the model and proofs. The modifications we study are to add omissive-asymetric faults, make the implementation time-triggered, and replace the majority vote with a mid-value selection.

@@

First, we briefly describe the algorithm, sketch our model of the algorithm within the framework presented in Section~\ref{sec:model}, then describe it's invariants.

\subsection{$\OMH(m)$ Algorithm}
$\OMH$ is a recursive algorithm that proceeds in rounds of communication. Here we give a recursive specification for $\OMH(m)$, parameterized by the number of rounds, $m$. Consider a finite set of nodes $N$. Distinguish one node as the \emph{general}, $g$, and the remaining nodes $L = N \setminus \{g\}$ as the \emph{lieutenants}. We assume the identity of any general or lieutenant cannot be spoofed. Broadcast communication proceeds in rounds. Denote any message that is detectably faulty (e.g., fails a CRC) or is absent, by $\ERR$. Additionally, in the algorithm, nodes report on values they have previously received. In doing so, nodes must differentiate \emph{reporting} $\ERR$ from an $\ERR$ itself. Let $R$ denote that an error is being reported. Finally, let $V$ be a special, designated value.

@@

\begin{itemize}
\item {\bf $\OM(0)$}: $g$ broadcasts a value to each lieutenant, and the lieutenants return the value received (or $\ERR$).
\item {\bf $\OM(m)$, $m > 0$}:
  \begin{enumerate}
  \item $g$ broadcasts a value to each lieutenant, $l$.
  \item\label{om:two} Let $l_v$ be the value received by $l \in L$ from $g$. Then for each $l$, execute $\OMH(m-1)$, assigning $l$ to be the general and $L \setminus \{l\}$ to be the lieutenants. $l$ sends $l_v$ or $R$, if $l_v = \ERR$.
  \item For each lieutenant $l \in L$, remove all $\ERR$ values received in Step~\ref{om:two} from executing $OHM(m-1)$. Compute the majority value over the remaining values, or $V$ if there is no majority. If the majority value is $R$, return $E$.
  \end{enumerate}
\end{itemize}

@@

$\OMH$ is designed to ensure \emph{validity} and \emph{agreement} properties under suitable hypotheses on the number and type of faults in the system. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. More formally, let the general broadcast $v \neq \ERR$, and let $l_i$ denote the output for lieutenant $i$:
%
\begin{equation}
  \tag{Validity}
    \forall \,i. \quad l_i = v
\end{equation}
%
Agreement states that each lieutenant outputs the same value. Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively:
%
\begin{equation}
  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j
\end{equation}
%

@@

\subsection{Model Sketch}\label{sec:sketch}

We have implemented $\OMH(1)$ (as well as many variants, described in Section~\ref{sec:modular}) in the Symbolic Analysis Laboratory~\cite{SAL} (SAL) language. The verification is also carried out in SAL, which contains a suite of model-checkers. In our work, we use infinite-state (SMT-based) $k$-induction~\cite{Moura_boundedmodel,Armando-et-al}.

We have chosen to follow Rushby~\cite{Rushby:OM1} in ``unrolling'' the communication among lieutenants into two sets of logical nodes: relays and receivers. A special node called the source takes the place of the general above. Thus, for example, in the case where 4 nodes execute $\OMH(1)$, our model consists of one source, three relays, and three receivers. We model communication through one-way, typed channels. The source broadcasts a message to each relay which, in turn, each broadcast their messages to all receivers.

Our SAL model defines 7 transition systems in total: \texttt{clock}, {\texttt{source}, \texttt{relay} (parametrized over an ID), \texttt{receiver} (parametrized over an ID), \texttt{abstractor}, \texttt{observer}, \texttt{abstract\_monitor}. The first four of these are composed asynchronously, in an intermediate system we label \texttt{system}, and share access to a global calendar consisting of event slots $(\text{message}, \text{time})$, one for each channel in the system. \texttt{clock} is responsible for updating a global variable \texttt{t} representing time according to the rules in Section~\ref{sec:calendar}. The asynchronous composition gives us a relaxed form of the original specification, but it is possible without much additional work to enforce timing constraints, see Section~\ref{sec:modular} "Time-Triggered Messaging". The remaining three systems are observers of the previously defined asynchronous system. The \texttt{observer} controls the state of \texttt{validity} and \texttt{agreement} variables, setting them to \texttt{FALSE} if, when the receivers have computed their vote, the properties do not hold. The \texttt{abstractor} and \texttt{abstract\_monitor} monitor the state of \texttt{system} and update the values of a collection of 6 abstract state predicates. These predicates express different increasing levels of progress of \texttt{system}: from the source sending its message, to the relays broadcasting, to the receivers voting on received values.

@@

\begin{figure}
  \centering
  \begin{tabular}{c|c||c|c|c|c|c|c|c|c|c|c|}
      \multicolumn{2-12}{c}{Receivers} \\
      \cline{2-12}
          &     & 1  &  2  &   3  &  4  &   5  &  6   &  7   &  8  &  9  &  10 \\
      \cline{2-12} \cline{2-12}
      \multirow{10}{*}{Relays} &
            1   & 7    & 9   & 12   & 15  & 21   & 25   & 32   & 40  & 54  & 74  \\
      \cline{2-12}
          & 2   & 17   & 14  & 21   & 30  & 42   & 53   & 74   & 99  & 144 & -   \\
      \cline{2-12}
          & 3   & 21   & 22  & 40   & 50  & 81   & 102  & 155  & 279 & -   & -   \\
      \cline{2-12}
          & 4   & 27   & 34  & 59   & 99  & 141  & 237  & 1114 & -   & -   & -   \\
      \cline{2-12}
          & 5   & 22   & 94  & 125  & 335 & \TO   & 1406 & -    & -   & -   & -   \\
      \cline{2-12}
          & 6   & 36   & 132 & 2966 & 844 & 2457 & -    & -    & -   & -   & -   \\
      \cline{2-12}
          & 7   & 83   & 487 & \TO   & \TO  & -    & -    & -    & -   & -   & -   \\
      \cline{2-12}
          & 8   & 298  & \TO  & \TO   & -   & -    & -    & -    & -   & -   & -   \\
      \cline{2-12}
          & 9   & 1428 & \TO  & -    & -   & -    & -    & -    & -   & -   & -   \\
      \cline{2-12}
          & 10   & \TO   & -   & -    & -   & -    & -    & -    & -   & -   & -   \\
      \cline{2-12}
  \end{tabular}
  \caption{Benchmark of full proof computation time for $\OMH(1)$ implementation. Times are in seconds with a timeout (\TO) limit of one hour. Dashes ('-') denote no benchmark was run.}
  \label{fig:benchmark}
\end{figure}

@@

As a point of comparison, we also benchmarked Rushby's model~\cite{Rushby:OM1} on the same machine, using the same environment. When the number of receivers was fixed at 2, the verification time for increasing numbers of relays was: $0.12 s$, $0.20 s$, $2.56 s$, $8.24 s$, $28.0 s$, $549 s$, \TO, \TO, \TO (compare to the numbers in the second column of Figure~\ref{fig:benchmark}). Empirical evidence suggests that our verification, while slower for small numbers of nodes, scales much better for larger numbers of nodes.

@@

\begin{figure}
  \centering
  \begin{tabular}{|p{2.25cm}|p{2.25cm}|p{2.25cm}|p{2.25cm}|p{2.25cm}|}
  \hline
                                  & Transition systems (7 total)  &Definitions\linebreak (58 total) & Invariants\linebreak (11 total) & Invariant classes (5 total) \\
  \hline \hline
  No Fault Model                  & receivers & 8 modified & 2 modified & ASM, faults \\
  \hline
  Omissive Asymmetric Faults       & none      & 4 modified & 2 modified & faults             \\
  \hline
  Time-Triggered Messaging        & source, relays, receivers & 3 new & 2 new, 3 modified & calendar, faults \\
  \hline
  Mid-Value Selection & receivers & 4 new, 3 modified & 2 modified & ASM, voting \\
  \hline
  \end{tabular}
  \caption{Refactoring effort for protocol modifications, measured by which portions of the model have to be modified.}
  \label{fig:effort}
\end{figure}

@@

\paragraph{Omissive Asymmetric Faults.}
Removing faults already described by the fault model is easy. Recall that in our model faults do not appear in the system specification and only operate on the calendar. Removing a fault from the system requires only setting the number of a particular kind of faults to zero in the maximum fault assumption.

@@

\paragraph{Time-Triggered Messaging.}
A \emph{time-triggered} distributed system is one in which nodes are independently clocked, but clock constraints (ensured by periodic clock synchronization) allow the model to appear as if it is executing synchronously~\cite{}.

@@

\paragraph{Mid-Value Selection.}
Our $\OMH(1)$ model leverages majority vote in order to tolerate faults. Another choice for the fault masking algorithm used here is mid-value selection (or median). This choice is more common in applications involving hardware, signal selection (where signal values cannot be relied on to exactly match even in the absence of faults), or cases where information about congruence is useful. To implement mid-value selection in our model, we allow messages sent to take values in $\mathbb{R}$ and the receiver transition system is modified in two ways. First, a second buffer is introduced which will hold the sorted contents of the main buffer once voting has commenced. Second, a mid-value select function is called on the sorted buffer and the result is stored as the receiver's vote. The only invariants needing modification were the ASM definition (to account for the values stored by the new buffer and the relation between it and the main buffer) and the voting invariant.

\subsection{Proof Effort Remarks}

The lemmas described in Section~\ref{sec:invariants} are constructed by-hand and represent multiple days of effort, but that effort includes both model and protocol construction and generalization as well as verification. The counterexamples returned by SAL are very useful for strengthening invariants, but tedious to analyze---a model with five relays and two receivers contains 90 state variables, and there are known counterexamples to models that size~\cite{csl-93-2}. Once we developed the ASM synchronously-composed observer, the verification effort was sped up considerably.
